---
title: "besthr - Generating Bootstrap Estimation Distributions of HR Data"
author: "Dan MacLean"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  github_document:
    html_preview: false
---

  <!-- badges: start -->
  [![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.3374507.svg)](https://doi.org/10.5281/zenodo.3374507)
  [![R-CMD-check](https://github.com/TeamMacLean/besthr/workflows/R-CMD-check/badge.svg)](https://github.com/TeamMacLean/besthr/actions)
  <!-- badges: end -->

## Synopsis

besthr is a package that creates plots showing scored HR experiments and plots of distribution of means of ranks of HR score from bootstrapping.


## Installation

You can install from CRAN in the usual way.

```{r, eval=FALSE}
install.packages("besthr")

# or for the dev version
#install.packages("devtools")
devtools::install_github("TeamMacLean/besthr")
```

## Citation

Please cite as 

> Dan MacLean. (2019). TeamMacLean/besthr: Initial Release (0.3.0). Zenodo. https://doi.org/10.5281/zenodo.3374507

## Simplest Use Case - Two Groups, No Replicates

With a data frame or similar object, use the `estimate()` function to get the bootstrap estimates of the ranked data.

`estimate()` has a basic function call as follows:

`estimate(data, score_column_name, group_column_name, control = control_group_name)`

The first argument after the 

```{r, fig.width=8, fig.height=6}
library(besthr)

hr_data_1_file <- system.file("extdata", "example-data-1.csv", package = "besthr")
hr_data_1 <- readr::read_csv(hr_data_1_file)
head(hr_data_1)

hr_est_1 <- estimate(hr_data_1, score, group, control = "A")
hr_est_1

plot(hr_est_1)
```


### Setting Options

You may select the group to set as the common reference control with `control`.

```{r, fig.width=8, fig.height=6}
estimate(hr_data_1, score, group, control = "B" ) %>%
  plot()
```

You may select the number of iterations of the bootstrap to perform with `nits` and the quantiles for the confidence interval with `low` and `high`.

```{r, fig.width=8, fig.height=6}
estimate(hr_data_1, score, group, control = "A", nits = 1000, low = 0.4, high = 0.6) %>%
  plot()
```
  
## Extended Use Case - Technical Replicates

You can extend the `estimate()` options to specify a third column in the data that contains technical replicate information, add the technical replicate column name after the sample column. Technical replicates are automatically merged using the `mean()` function before ranking.

```{r, fig.width=8, fig.height=6}

hr_data_3_file <- system.file("extdata", "example-data-3.csv", package = "besthr")
hr_data_3 <- readr::read_csv(hr_data_3_file)
head(hr_data_3)

hr_est_3 <- estimate(hr_data_3, score, sample, rep, control = "A")

hr_est_3

plot(hr_est_3)

```

### Alternate Plot Options

In the case where you have use technical replicates and want to see those plotted you can use an extra plot option `which`. Set `which` to `just_data` if you wish the left panel of the plot to show all data without ranking. This will only work if you have technical replicates.

```{r, fig.width=8, fig.height=6}

hr_est_3 %>%
  plot(which = "just_data")
```

## Built-in Themes and Color Palettes

besthr includes built-in themes and colorblind-safe color palettes that can be applied directly through the `plot()` function.

### Theme Options

Use the `theme` parameter to change the overall visual style:

- `"classic"` (default) - The original besthr appearance
- `"modern"` - A cleaner, contemporary style with refined typography and grid

```{r, fig.width=8, fig.height=6}
# Classic theme (default - same as before)
plot(hr_est_1, theme = "classic")

# Modern theme
plot(hr_est_1, theme = "modern")
```

### Color Palette Options

Use the `colors` parameter to change the color palette:

- `"default"` - Original besthr colors
- `"okabe_ito"` - Colorblind-safe Okabe-Ito palette
- `"viridis"` - Viridis color scale

```{r, fig.width=8, fig.height=6}
# Colorblind-safe palette
plot(hr_est_1, colors = "okabe_ito")

# Viridis palette
plot(hr_est_1, colors = "viridis")
```

### Combining Theme and Colors

You can combine both options for a fully customized look:

```{r, fig.width=8, fig.height=6}
# Modern theme with colorblind-safe colors
plot(hr_est_1, theme = "modern", colors = "okabe_ito")
```

### Using besthr Palettes Directly

The color palettes can also be used directly in your own ggplot2 code:
```{r}
# Get palette colors
besthr_palette("okabe_ito", n = 4)

# Available palettes
besthr_palette("default", n = 3)
besthr_palette("viridis", n = 3)
```

## Styling Plots

You can style plots to your own taste. The object returned from `plot()` is a `patchwork` [https://patchwork.data-imaginist.com/](https://patchwork.data-imaginist.com/) object that composes two separate plots, the dot plot and the bootstrap percentile plot, which are themselves `ggplot` objects. So you can use a mixture of `patchwork` annotations functions for whole plot labels and `ggplot` themes for individual elements. 

### Adding annotations.

You can use the `patchwork` `plot_annotation()` function to add titles

```{r, fig.width=8, fig.height=6}
library(patchwork)

p <- plot(hr_est_1)

p + plot_annotation(title = 'A stylish besthr plot', 
                    subtitle = "better than ever", 
                    caption = 'Though this example is not meaningful')
p


```


### Targetting a subplot to make theme changes

You can change the style of the individual plot elements using subsetting syntax `[[]]` . The dot plot can be addressed within the `patchwork` object using index 1 within the `patchwork` object `p[[1]]`, and the percentile plot using `p[[2]]`. You must add to the existing subplot then assign the result back to see the difference in the plot. Here's an example that uses `theme()` to restyle the y-axis text of the dot plot

```{r, fig.width=8, fig.height=6}
library(ggplot2)
p[[1]] <- p[[1]] + theme(axis.title.y = element_text(family = "Times", colour="blue", size=24))
p
```


### Changing the scale colours of a subplot

You can change the colours used by the scales in the same way using the `scale` functions, though as the type of scale is different for the dot plot and bootstrap plot you will need to apply a different scale for each.

For the dot plot, use a discrete scale e.g `scale_colour_manual()`, `scale_colour_viridis_d()` or `scale_colour_brewer(type = "qual")`

```{r, fig.width=8, fig.height=6}
p[[1]] <- p[[1]] + scale_colour_manual(values = c("blue", "#440000"))
p

p[[1]] <- p[[1]] + scale_colour_viridis_d()
p

p[[1]] <- p[[1]] + scale_colour_brewer(type="qual", palette="Accent")
p
```

For the percentile plot, use only `scale_colour_manual()` with specified colours. Annoyingly, this rewrites the other values associated with the scale each time, so you'll need to replace those.

```{r, fig.width=8, fig.height=6}
p[[2]] <- p[[2]] + scale_fill_manual(
  values = c("blue", "pink", "yellow"),
  name = "bootstrap percentile", labels=c("lower", "non-significant", "higher"),
  guide = guide_legend(reverse=TRUE)
  )
p
```

## Alternative Visualizations

### Forest Plot

For publication-style forest plots showing point estimates with confidence intervals:

```{r, fig.width=8, fig.height=4}
# Basic forest plot
plot_forest(hr_est_1)

# With theme options
plot_forest(hr_est_1, theme = "modern", colors = "okabe_ito")
```

### Raincloud Plot

For a combined view of raw data, density, and summary statistics:

```{r, fig.width=8, fig.height=6}
plot_raincloud(hr_est_1)
```

## Significance and Effect Size Annotations

You can add statistical annotations to your plots to highlight significant results.

### Significance Stars

Add significance stars to groups where the bootstrap confidence interval does not overlap the control mean:

```{r, fig.width=8, fig.height=6}
# Create data with a clear difference for demonstration
d_sig <- data.frame(
  score = c(rep(2, 10), rep(8, 10)),
  group = rep(c("A", "B"), each = 10)
)
hr_sig <- estimate(d_sig, score, group, control = "A", nits = 500)

# Plot with significance annotation
plot(hr_sig, show_significance = TRUE)
```

### Effect Size Annotation

Display effect size (difference from control) with confidence intervals:

```{r, fig.width=8, fig.height=6}
plot(hr_sig, show_effect_size = TRUE)
```

### Computing Statistics Directly

You can also access the significance and effect size calculations directly:

```{r}
# Compute significance
compute_significance(hr_est_1)

# Compute effect sizes
compute_effect_size(hr_est_1)
```

## Summary Tables

Generate publication-ready summary tables with `besthr_table()`:

```{r}
# Default tibble format
besthr_table(hr_est_1)

# With significance stars
besthr_table(hr_sig, include_significance = TRUE)
```

### Export Formats

Generate tables in various formats for publication:

```{r}
# Markdown format
besthr_table(hr_est_1, format = "markdown")
```

```{r, eval=FALSE}
# HTML format
besthr_table(hr_est_1, format = "html")

# LaTeX format
besthr_table(hr_est_1, format = "latex")
```

## Publication Export

Save your plots directly to publication-quality files:

```{r, eval=FALSE}
# Save to PNG (default 300 DPI)
save_besthr(hr_est_1, "figure1.png")

# Save to PDF
save_besthr(hr_est_1, "figure1.pdf", width = 10, height = 8)

# Save forest plot
save_besthr(hr_est_1, "forest.png", type = "forest")

# Save raincloud plot
save_besthr(hr_est_1, "raincloud.png", type = "raincloud")

# With custom options
save_besthr(hr_est_1, "figure1.png",
            theme = "modern",
            colors = "okabe_ito",
            width = 10,
            height = 6,
            dpi = 600)
```

Supported formats: PNG, PDF, SVG, TIFF, JPEG

