---
title: "besthr - Generating Bootstrap Estimation Distributions of HR Data"
author: "Dan MacLean"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  github_document:
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.path = "figures/")
```

  <!-- badges: start -->
  [![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.3374507.svg)](https://doi.org/10.5281/zenodo.3374507)
  [![R-CMD-check](https://github.com/TeamMacLean/besthr/workflows/R-CMD-check/badge.svg)](https://github.com/TeamMacLean/besthr/actions)
  [![codecov](https://codecov.io/gh/TeamMacLean/besthr/branch/develop/graph/badge.svg)](https://codecov.io/gh/TeamMacLean/besthr)
  <!-- badges: end -->

## Synopsis

besthr is a package that creates plots showing scored HR experiments and plots of distribution of means of ranks of HR score from bootstrapping.


## Installation

You can install from CRAN in the usual way.

```{r, eval=FALSE}
install.packages("besthr")

# or for the dev version
#install.packages("devtools")
devtools::install_github("TeamMacLean/besthr")
```

## Citation

Please cite as 

> Dan MacLean. (2019). TeamMacLean/besthr: Initial Release (0.3.0). Zenodo. https://doi.org/10.5281/zenodo.3374507

## Simplest Use Case - Two Groups, No Replicates

With a data frame or similar object, use the `estimate()` function to get the bootstrap estimates of the ranked data.

`estimate()` has a basic function call as follows:

`estimate(data, score_column_name, group_column_name, control = control_group_name)`

The first argument after the 

```{r, fig.width=8, fig.height=6}
library(besthr)

hr_data_1_file <- system.file("extdata", "example-data-1.csv", package = "besthr")
hr_data_1 <- readr::read_csv(hr_data_1_file)
head(hr_data_1)

hr_est_1 <- estimate(hr_data_1, score, group, control = "A")
hr_est_1

plot(hr_est_1)
```


### Setting Options

You may select the group to set as the common reference control with `control`.

```{r, fig.width=8, fig.height=6}
estimate(hr_data_1, score, group, control = "B" ) %>%
  plot()
```

You may select the number of iterations of the bootstrap to perform with `nits` and the quantiles for the confidence interval with `low` and `high`.

```{r, fig.width=8, fig.height=6}
estimate(hr_data_1, score, group, control = "A", nits = 1000, low = 0.4, high = 0.6) %>%
  plot()
```
  
## Extended Use Case - Technical Replicates

You can extend the `estimate()` options to specify a third column in the data that contains technical replicate information, add the technical replicate column name after the sample column. Technical replicates are automatically merged using the `mean()` function before ranking.

```{r, fig.width=8, fig.height=6}

hr_data_3_file <- system.file("extdata", "example-data-3.csv", package = "besthr")
hr_data_3 <- readr::read_csv(hr_data_3_file)
head(hr_data_3)

hr_est_3 <- estimate(hr_data_3, score, sample, rep, control = "A")

hr_est_3

plot(hr_est_3)

```

### Alternate Plot Options

In the case where you have use technical replicates and want to see those plotted you can use an extra plot option `which`. Set `which` to `just_data` if you wish the left panel of the plot to show all data without ranking. This will only work if you have technical replicates.

```{r, fig.width=8, fig.height=6}

hr_est_3 %>%
  plot(which = "just_data")
```

## Built-in Themes and Color Palettes

besthr includes built-in themes and colorblind-safe color palettes that can be applied directly through the `plot()` function.

### Theme Options

Use the `theme` parameter to change the overall visual style:

- `"classic"` (default) - The original besthr appearance
- `"modern"` - A cleaner, contemporary style with refined typography and grid

```{r, fig.width=8, fig.height=6}
# Classic theme (default - same as before)
plot(hr_est_1, theme = "classic")

# Modern theme
plot(hr_est_1, theme = "modern")
```

### Color Palette Options

Use the `colors` parameter to change the color palette:

- `"default"` - Original besthr colors
- `"okabe_ito"` - Colorblind-safe Okabe-Ito palette
- `"viridis"` - Viridis color scale

```{r, fig.width=8, fig.height=6}
# Colorblind-safe palette
plot(hr_est_1, colors = "okabe_ito")

# Viridis palette
plot(hr_est_1, colors = "viridis")
```

### Combining Theme and Colors

You can combine both options for a fully customized look:

```{r, fig.width=8, fig.height=6}
# Modern theme with colorblind-safe colors
plot(hr_est_1, theme = "modern", colors = "okabe_ito")
```

### Using besthr Palettes Directly

The color palettes can also be used directly in your own ggplot2 code:
```{r}
# Get palette colors
besthr_palette("okabe_ito", n = 4)

# Available palettes
besthr_palette("default", n = 3)
besthr_palette("viridis", n = 3)
```

## Styling Plots

### Recommended: Use Built-in Themes and Colors

The easiest way to style your plots is using the `theme` and `colors` parameters:

```{r, fig.width=8, fig.height=6}
# Modern look with colorblind-safe colors (this is the default)
plot(hr_est_1, theme = "modern", colors = "okabe_ito")

# Classic appearance
plot(hr_est_1, theme = "classic", colors = "default")

# Viridis color scheme
plot(hr_est_1, colors = "viridis")
```

### Adding Titles and Annotations

The plot object is a `patchwork` composition. You can add titles using `plot_annotation()`:

```{r, fig.width=8, fig.height=6}
library(patchwork)

p <- plot(hr_est_1)

p + plot_annotation(
  title = 'HR Score Analysis',
  subtitle = "Control vs Treatment",
  caption = 'Generated with besthr'
)
```

## Raincloud Plot

For a combined view of raw data points with summary statistics, use `plot_raincloud()`:

```{r, fig.width=8, fig.height=6}
plot_raincloud(hr_est_1)
```

## Significance and Effect Size Annotations

You can add statistical annotations to your plots to highlight significant results.

```{r}
# Create example data with 3 groups and realistic variation
set.seed(42)
d_effect <- data.frame(
  score = c(
    sample(1:4, 12, replace = TRUE),   # Group A: low scores (control)
    sample(4:8, 12, replace = TRUE),   # Group B: medium-high scores
    sample(6:10, 12, replace = TRUE)   # Group C: high scores
  ),
  group = rep(c("A", "B", "C"), each = 12)
)
hr_effect <- estimate(d_effect, score, group, control = "A", nits = 1000)
```

### Significance Stars

Add significance stars to groups where the bootstrap confidence interval does not overlap the control mean:

```{r, fig.width=8, fig.height=6}
plot(hr_effect, show_significance = TRUE)
```

### Effect Size Annotation

Display effect size (difference from control) with confidence intervals:

```{r, fig.width=8, fig.height=6}
plot(hr_effect, show_effect_size = TRUE)
```

### Computing Statistics Directly

You can also access the significance and effect size calculations directly:

```{r}
# Compute significance
compute_significance(hr_est_1)

# Compute effect sizes
compute_effect_size(hr_est_1)
```

## Summary Tables

Generate publication-ready summary tables with `besthr_table()`:

```{r}
# Default tibble format
besthr_table(hr_effect)

# With significance stars
besthr_table(hr_effect, include_significance = TRUE)
```

### Export Formats

Generate tables in various formats for publication:

```{r}
# Markdown format
besthr_table(hr_est_1, format = "markdown")
```

```{r, eval=FALSE}
# HTML format
besthr_table(hr_est_1, format = "html")

# LaTeX format
besthr_table(hr_est_1, format = "latex")
```

## Publication Export

Save your plots directly to publication-quality files:

```{r, eval=FALSE}
# Save to PNG (default 300 DPI)
save_besthr(hr_est_1, "figure1.png")

# Save to PDF
save_besthr(hr_est_1, "figure1.pdf", width = 10, height = 8)

# Save raincloud plot
save_besthr(hr_est_1, "raincloud.png", type = "raincloud")

# With custom options
save_besthr(hr_est_1, "figure1.png",
            theme = "modern",
            colors = "okabe_ito",
            width = 10,
            height = 6,
            dpi = 600)
```

Supported formats: PNG, PDF, SVG, TIFF, JPEG

