---
title: "besthr - Generating Bootstrap Estimation Distributions of HR Data"
author: "Dan MacLean"
date: "12/06/2019"
output: 
  github_document:
    html_preview: false
---

  <!-- badges: start -->
  ![Travis build status](https://travis-ci.org/TeamMacLean/besthr.svg?branch=master)

  <!-- badges: end -->

## Synopsis

besthr is a package that creates plots showing scored HR experiments and plots of distribution of means of ranks of HR score from bootstrapping.

```{r}
library(besthr)
hr_data_1_file <- system.file("extdata", "example-data-1.csv", package = "besthr")
hr_data_1 <- readr::read_csv(hr_data_1_file)
head(hr_data_1)

hr_est_1 <- estimate(hr_data_1, score, group, control = "A")
hr_est_1

plot(hr_est_1)

```

## Installation

```{r, eval=FALSE}
devtools::install_github("TeamMacLean/besthr")
```

## Usage 

### Simplest Use Case - Two Groups, No Replicates

With a data frame or similar object, use the `estimate()` function to get the bootstrap estimates of the ranked data.

`estimate()` has a basic function call as follows:

`estimate(data, score_column_name, group_column_name, control = control_group_name)`

The first argument after the data frame is the title of the score column, the second the title of the group (or sample) column. 

```{r}
library(besthr)

hr_data_1_file <- system.file("extdata", "example-data-1.csv", package = "besthr")
hr_data_1 <- readr::read_csv(hr_data_1_file)
head(hr_data_1)

hr_est_1 <- estimate(hr_data_1, score, group, control = "A")
hr_est_1

plot(hr_est_1)
```

#### Setting Options

You may select the group to set as the common reference control with `control`.

```{r}
estimate(hr_data_1, score, group, control = "B" ) %>%
  plot()
```

You may select the number of iterations of the bootstrap to perform with `nits` and the quantiles for the confidence interval with `low` and `high`.

```{r}
estimate(hr_data_1, score, group, control = "A", nits = 1000, low = 0.4, high = 0.6) %>%
  plot()
```
  
### Extended Use Case - Technical Replicates

You can extend the `estimate()` options to specify a third column in the data that contains technical replicate information, add the technical replicate column name after the sample column. Technical replicates are automatically merged using the `mean()` function before ranking.

```{r}

hr_data_3_file <- system.file("extdata", "example-data-3.csv", package = "besthr")
hr_data_3 <- readr::read_csv(hr_data_3_file)
head(hr_data_3)

hr_est_3 <- estimate(hr_data_3, score, sample, rep, control = "A")

hr_est_3

plot(hr_est_3)

```

#### Alternate Plot Options

In the case where you have use technical replicates and want to see those plotted you can use an extra plot option `which`. Set `which` to `just_data` if you wish the left panel of the plot to show all data without ranking. This will only work if you have technical replicates.

```{r}

hr_est_3 %>% 
  plot(which = "just_data")
```
