---
title: "basic-use"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{basic-use}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Citation

Please cite as 

> Dan MacLean. (2019). TeamMacLean/besthr: Initial Release (0.3.0). Zenodo. https://doi.org/10.5281/zenodo.3374507

## Simplest Use Case - Two Groups, No Replicates

With a data frame or similar object, use the `estimate()` function to get the bootstrap estimates of the ranked data.

`estimate()` has a basic function call as follows:

`estimate(data, score_column_name, group_column_name, control = control_group_name)`

The first argument after the 

```{r, fig.width=8, fig.height=6}
library(besthr)

hr_data_1_file <- system.file("extdata", "example-data-1.csv", package = "besthr")
hr_data_1 <- readr::read_csv(hr_data_1_file)
head(hr_data_1)

hr_est_1 <- estimate(hr_data_1, score, group, control = "A")
hr_est_1

plot(hr_est_1)
```


### Setting Options

You may select the group to set as the common reference control with `control`.

```{r, fig.width=8, fig.height=6}
estimate(hr_data_1, score, group, control = "B" ) %>%
  plot()
```

You may select the number of iterations of the bootstrap to perform with `nits` and the quantiles for the confidence interval with `low` and `high`.

```{r, fig.width=8, fig.height=6}
estimate(hr_data_1, score, group, control = "A", nits = 1000, low = 0.4, high = 0.6) %>%
  plot()
```
  
## Extended Use Case - Technical Replicates

You can extend the `estimate()` options to specify a third column in the data that contains technical replicate information, add the technical replicate column name after the sample column. Technical replicates are automatically merged using the `mean()` function before ranking.

```{r, fig.width=8, fig.height=6}

hr_data_3_file <- system.file("extdata", "example-data-3.csv", package = "besthr")
hr_data_3 <- readr::read_csv(hr_data_3_file)
head(hr_data_3)

hr_est_3 <- estimate(hr_data_3, score, sample, rep, control = "A")

hr_est_3

plot(hr_est_3)

```

### Alternate Plot Options

In the case where you have use technical replicates and want to see those plotted you can use an extra plot option `which`. Set `which` to `just_data` if you wish the left panel of the plot to show all data without ranking. This will only work if you have technical replicates.

```{r, fig.width=8, fig.height=6}

hr_est_3 %>%
  plot(which = "just_data")
```

## Built-in Themes and Color Palettes

besthr uses a modern, colorblind-safe appearance by default. You can customize the look using themes and color palettes.

### Theme Options

Use the `theme` parameter to change the overall visual style:

- `"modern"` (default) - A clean, contemporary style with refined typography
- `"classic"` - The original besthr appearance (for backward compatibility)

```{r, fig.width=8, fig.height=6}
# Modern theme (default)
plot(hr_est_1)

# Classic theme (original style)
plot(hr_est_1, theme = "classic")
```

### Color Palette Options

Use the `colors` parameter to change the color palette:

- `"okabe_ito"` (default) - Colorblind-safe Okabe-Ito palette
- `"default"` - Original besthr colors (for backward compatibility)
- `"viridis"` - Viridis color scale

```{r, fig.width=8, fig.height=6}
# Default is already colorblind-safe
plot(hr_est_1)

# Original colors
plot(hr_est_1, colors = "default")

# Viridis palette
plot(hr_est_1, colors = "viridis")
```

### Quick Style Presets

For easy customization without understanding all options, use preset styles:

```{r, fig.width=8, fig.height=6}
# Publication-ready style
plot(hr_est_1, config = besthr_style("publication"))

# Presentation style (larger elements)
plot(hr_est_1, config = besthr_style("presentation"))

# Density style (gradient density instead of points)
plot(hr_est_1, config = besthr_style("density"))

# See all available styles
list_besthr_styles()
```

### Using besthr Palettes Directly

The color palettes can also be used directly in your own ggplot2 code:
```{r}
# Get palette colors
besthr_palette("okabe_ito", n = 4)

# Available palettes
besthr_palette("default", n = 3)
besthr_palette("viridis", n = 3)
```

## Styling Plots

You can style plots to your own taste. The object returned from `plot()` is a `patchwork` [https://patchwork.data-imaginist.com/](https://patchwork.data-imaginist.com/) object that composes two separate plots, the dot plot and the bootstrap percentile plot, which are themselves `ggplot` objects. So you can use a mixture of `patchwork` annotations functions for whole plot labels and `ggplot` themes for individual elements. 

### Adding annotations.

You can use the `patchwork` `plot_annotation()` function to add titles

```{r, fig.width=8, fig.height=6}
library(patchwork)

p <- plot(hr_est_1)

p + plot_annotation(title = 'A stylish besthr plot', 
                    subtitle = "better than ever", 
                    caption = 'Though this example is not meaningful')
p


```


### Targetting a subplot to make theme changes

You can change the style of the individual plot elements using subsetting syntax `[[]]` . The dot plot can be addressed within the `patchwork` object using index 1 within the `patchwork` object `p[[1]]`, and the percentile plot using `p[[2]]`. You must add to the existing subplot then assign the result back to see the difference in the plot. Here's an example that uses `theme()` to restyle the y-axis text of the dot plot

```{r, fig.width=8, fig.height=6}
library(ggplot2)
p[[1]] <- p[[1]] + theme(axis.title.y = element_text(family = "Times", colour="blue", size=24))
p
```


### Changing the scale colours of a subplot

You can change the colours used by the scales in the same way using the `scale` functions, though as the type of scale is different for the dot plot and bootstrap plot you will need to apply a different scale for each.

For the dot plot, use a discrete scale e.g `scale_colour_manual()`, `scale_colour_viridis_d()` or `scale_colour_brewer(type = "qual")`

```{r, fig.width=8, fig.height=6}
p[[1]] <- p[[1]] + scale_colour_manual(values = c("blue", "#440000"))
p

p[[1]] <- p[[1]] + scale_colour_viridis_d()
p

p[[1]] <- p[[1]] + scale_colour_brewer(type="qual", palette="Accent")
p
```

For the percentile plot, use only `scale_colour_manual()` with specified colours. Annoyingly, this rewrites the other values associated with the scale each time, so you'll need to replace those.

```{r, fig.width=8, fig.height=6}
p[[2]] <- p[[2]] + scale_fill_manual(
  values = c("blue", "pink", "yellow"),
  name = "bootstrap percentile", labels=c("lower", "non-significant", "higher"),
  guide = guide_legend(reverse=TRUE)
  )
p
```

## Advanced Configuration

For users who need fine-grained control over plot appearance, besthr provides a configuration system.

### Using besthr_plot_config

The `besthr_plot_config()` function creates a configuration object that controls all aspects of plot appearance:

```{r}
# Create a custom configuration
cfg <- besthr_plot_config(
  panel_widths = c(2, 1),        # Make data panel wider than bootstrap panel
  point_size_range = c(3, 10),   # Larger points
  point_alpha = 0.9,             # More opaque points
  mean_line_width = 1.5,         # Thicker mean lines
  theme_style = "modern",
  color_palette = "okabe_ito"
)

print(cfg)
```

```{r, fig.width=8, fig.height=6}
# Use the configuration in plot
plot(hr_est_1, config = cfg)
```

You can update an existing configuration with `update_config()`:

```{r, fig.width=8, fig.height=6}
# Modify just one setting
cfg2 <- update_config(cfg, panel_widths = c(1, 2))
plot(hr_est_1, config = cfg2)
```

### Configuration Options

| Parameter | Default | Description |
|-----------|---------|-------------|
| `panel_widths` | `c(1, 1)` | Relative widths of observation and bootstrap panels |
| `y_limits` | `NULL` (auto) | Fixed y-axis limits, or NULL for automatic |
| `y_expand` | `0.05` | Proportional expansion of y-axis limits |
| `point_size_range` | `c(2, 8)` | Min/max point sizes based on count |
| `point_alpha` | `0.8` | Point transparency (0-1) |
| `mean_line_type` | `3` | Line type for mean indicators |
| `mean_line_width` | `1` | Line width for mean indicators |
| `density_alpha` | `0.7` | Bootstrap density transparency |
| `density_style` | `"points"` | Density display: "points" (jittered bootstrap), "gradient", or "solid" |
| `theme_style` | `"modern"` | Theme: "classic" or "modern" |
| `color_palette` | `"okabe_ito"` | Colors: "default", "okabe_ito", or "viridis" |

## Building Custom Plots

For maximum flexibility, you can build plots from individual components.

### Using besthr_data_view

The `besthr_data_view()` function extracts all plotting data from an hrest object with unified axis limits:

```{r}
# Create a data view
dv <- besthr_data_view(hr_est_1)
print(dv)

# Access the data
head(dv$ranked)
head(dv$bootstrap)
dv$rank_limits
```

### Using Panel Builders

You can build individual panels and compose them yourself:

```{r, fig.width=10, fig.height=5}
library(patchwork)

# Create panels
cfg <- besthr_plot_config(theme_style = "modern", color_palette = "okabe_ito")
dv <- besthr_data_view(hr_est_1, cfg)

p1 <- build_observation_panel(dv, cfg, "rank_simulation")
p2 <- build_bootstrap_panel(dv, cfg)

# Custom composition with different widths
p1 + p2 + plot_layout(widths = c(3, 1))
```

## Alternative Visualizations

### Raincloud Plots

The `plot_raincloud()` function provides an alternative visualization showing jittered data points with confidence intervals:

```{r, fig.width=6, fig.height=5}
plot_raincloud(hr_est_1)
```

```{r, fig.width=6, fig.height=5}
# With modern styling
plot_raincloud(hr_est_1, theme = "modern", colors = "okabe_ito")
```

### Bootstrap Raincloud

The `plot_bootstrap_raincloud()` function shows the bootstrap distribution as jittered points:

```{r, fig.width=6, fig.height=5}
plot_bootstrap_raincloud(hr_est_1)
```

## CI Color Customization

The `derive_ci_colors()` function generates confidence interval colors that harmonize with the selected palette:

```{r}
# Default CI colors
derive_ci_colors("default", "classic")

# Okabe-Ito derived CI colors
derive_ci_colors("okabe_ito", "modern")

# Viridis derived CI colors
derive_ci_colors("viridis", "classic")
```


